\batcave: Calling somatic mutations with a\\tumor- and site-specific prior

### 2019-10-08 bioRxiv_v1.tex
Version for bioRxiv

\documentclass[a4,center,fleqn]{NAR}
\bibliographystyle{NAR-natbib}
\usepackage{color}
\usepackage{xfrac}

\usepackage{xspace}
\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
\newcommand{\batcave}{BATCAVE\xspace}
\newcommand{\beginsupplement}{%
        \clearpage
        \onecolumn
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Enter dates of publication
\copyrightyear{2008}
\pubdate{31 July 2009}
\pubyear{2009}
\jvolume{37}
\jissue{12}

\articlesubtype{Genomics and Bioinformatics}

@@

\title{\batcave: Calling somatic mutations with a tumor- and site-specific prior}

\author{%
Brian K. Mannakee\,$^{1}$ and
Ryan N. Gutenkunst\,$^{2}$%
\footnote{To whom correspondence should be addressed.
Email: rgutenk@email.arizona.edu}}

\address{%
$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
and
$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

\history{%
Received January 1, 2009;
Revised February 1, 2009;
Accepted March 1, 2009}

@@

\section{Introduction}

@@

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka2 \cite{Kim2018}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
\centering
  \includegraphics{figures/real_signatures_only.pdf}
  \caption{Real tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  \textbf{(C)}\&\textbf{(D)} The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

\section{MATERIALS AND METHODS}
\subsection{Somatic variant calling probability model}

@@

\subsection{Site-specific prior probability of mutation}
The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
But $\mathrm{p}(m,M)$ is not uniform across the genome.
Rather it depends on the local genomic context $C$, so its full form is $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' theorem to further decompose as 
\begin{equation}
  \label{eqn:full_prior}
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
\end{equation}
We next show how to estimate the quantities in Eq.~\ref{eqn:full_prior}.

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

\subsection{Likelihood function}
The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq. \ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all analyses in this paper, so we have
\begin{equation}
  \label{eqn:tlod}
    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
\end{equation}
The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
\begin{equation}
  \label{eqn:computed_posterior}
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
\end{equation}
Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
\begin{equation}  \label{eqn:mutect_posterior}
  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
\end{equation}

\subsection{Implementation}
We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

\begin{figure}
\centering
  \includegraphics{figures/sim_signatures_only.pdf}
  \caption{Simulated tumor mutation profiles.
As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  \textbf{(A)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(B)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(C)} Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}

\subsection{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp 
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.\\
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

\section{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.

@@

\subsection{Tests using simulated data}

@@

  \begin{figure*}
\centering
    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
    \textbf{(A)} Precision-recall curves and \textbf{(B)} receiver operating characteristic curves for different mutation profiles.
    \textbf{(C)} and \textbf{(D)} Calibration plots. Shaded regions show distributions of posterior probabilities for true positive variants, and smooth lines show loess-smoothed relationships, from which the Integrated Calibration Index is calculated. For a perfectly calibrated caller, those curves would match the dashed y=x line. 
    \textbf{(E)} and \textbf{(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
}
  \label{NAR-wes_fig}
  \end{figure*}

>>>>

\documentclass[a4,center,fleqn]{article}
\bibliographystyle{NAR-natbib}
\usepackage{color}
\usepackage{xfrac}

\usepackage{xspace}
\newcommand{\rngcomment}[1]{{\color{red}RNG: #1}}
\newcommand{\bkmcomment}[1]{{\color{blue}BKM: #1}}
\newcommand{\batcave}{BATCAVE\xspace}
\newcommand{\beginsupplement}{%
        \clearpage
        \onecolumn
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Enter dates of publication
%\copyrightyear{2008}
%\pubdate{31 July 2009}
%\pubyear{2009}
%\jvolume{37}
%\jissue{12}

%\articlesubtype{Genomics and Bioinformatics}

\usepackage{fullpage}
\usepackage{amsmath}

@@

\title{\batcave: Calling somatic mutations with a\\tumor- and site-specific prior}

\author{%
Brian K. Mannakee$^{1}$ and
Ryan N. Gutenkunst$^{2}$%
\footnote{To whom correspondence should be addressed.
Email: rgutenk@email.arizona.edu}}

%\address{%
%$^{1}$Mel and Enid Zuckerman College of Public Health, University of Arizona, Tucson AZ
%and
%$^{2}$Department of Molecular and Cellular Biology, University of Arizona, Tucson AZ}
% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

%\history{%
%Received January 1, 2009;
%Revised February 1, 2009;
%Accepted March 1, 2009}

@@

\section*{Introduction}

@@

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1~\cite{Cibulskis2013} and Varscan2~\cite{Koboldt2012}, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \cite{Garrison2012},
joint genotype analysis (SomaticSniper \cite{Larson2012}, JointSNVMix2 \cite{Roth2012a}, Seurat \cite{Christoforides2013}, CaVEMan \cite{Jones2016}, and MuClone \cite{Dorri2019}), allele frequency-based analysis (Strelka \cite{Saunders2012}, LoFreq \cite{Wilm2012}, EBCall \cite{Shiraishi2013b}, deepSNV \cite{Gerstung2012}, LoLoPicker \cite{Carrot-Zhang2017}, and MuSE \cite{Fan2016}), and ensemble and deep learning methods (MutationSeq \cite{Ding2012}, BAYSIC \cite{Cantarel2014}, SomaticSeq \cite{Fang2015}, and SNooPer \cite{Spinella2016}).
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
If fact, even the processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \cite{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, the mutation profiles differ substantially among the three breast tumors illustrated in Figure~\ref{NAR-realsigfig}B-D.
%Figure \ref{NAR-sigfig} {a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
\centering
  \includegraphics{figures/real_signatures_only.pdf}
  \caption{Real tumor mutation profiles.
  In each panel, the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  (A) The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  (B) The observed mutation profile of a breast tumor used in our real data analysis \cite{Shi2018}.
  (C)\&(D) The observed mutation profiles of two additional breast tumors \cite{Alexandrov2019}.
   }
  \label{NAR-realsigfig}
 \end{figure}

\section*{MATERIALS AND METHODS}
\subsection*{Somatic variant calling probability model}

@@

\subsection*{Site-specific prior probability of mutation}
The probability that we have denoted $\mathrm{p}(m)$ in Eq.~\ref{eqn:2} is more precisely the joint probability that a mutation has occurred $M$ and that it was to allele $m$, which we denote $\mathrm{p}(m,M)$.
But $\mathrm{p}(m,M)$ is not uniform across the genome.
Rather it depends on the local genomic context $C$, so its full form is $\mathrm{p}(m,M | C)$ \cite{Buisson2019}.
Assuming that $m$ and $M$ are independent conditional on the genomic context, $\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C)$, which we can use Bayes' theorem to further decompose as 
\begin{equation}
  \label{eqn:full_prior}
  \mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C)}.
\end{equation}
We next show how to estimate the quantities in Eq.~\ref{eqn:full_prior}.

\subsection*{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the trinucleotide context, which is known to strongly affect the prior probability of single-nucleotide mutation \cite{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The trinucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot 4$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \mathrm{p}(\mathrm{S}_{m,c}) = \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Fig.~\ref{NAR-realsigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

@@

\subsection*{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
%A value 0.12 for $f_{min}$ at lower sequencing depth is suggested in Williams et al., 2016~\cite{Williams2016}.
%Equation \ref{eqn:mut_rate} then allows us to estimate $\mathrm{p}(M) = \mu$, and we have a complete tumor-specific specification of the site-specific prior probability of mutation.

\subsection*{Likelihood function}
The current implementation of \batcave builds on MuTect, because MuTect reports the log ratio of the likelihood functions for the null and alternative hypotheses (Eq. \ref{eqn:hypothesis}) as \textrm{TLOD} (MuTect1) or \textrm{t\_lod\_fstar} (MuTect2).
We used MuTect 1.1.7 for all analyses in this paper, so we have
\begin{equation}
  \label{eqn:tlod}
    \mathrm{TLOD} = \mathrm{log}_{10}\left(\frac{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = \hat{f}}(x_i)}{\prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu = 0}(x_i)}\right).
\end{equation}
The log posterior odds is the log likelihood ratio (\textrm{TLOD}) plus the log prior odds, so the posterior odds in favor of the alternate hypothesis for a given substitution type is
\begin{equation}
  \label{eqn:computed_posterior}
  \frac{\mathrm{P}(m,\nu = \hat{f})}{1 - \mathrm{P}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} + \mathrm{logit}_{10}(\mathrm{p}(\mathrm{S}_{m,c}))}.
\end{equation}
Here $\mathrm{p}(\mathrm{S}_{m,c})$ is the prior probability of a substitution of type $\mathrm{S}_{m,c}$, as described in Eq.~\ref{eqn:detailed_prior} and specified in Eq.~\ref{eqn:post_pred}-\ref{eqn:mut_rate}.
When comparing our posterior odds to those of MuTect, we assume a uniform per-base probability of mutation of $3\cdot10^{-6}$ \cite{Cibulskis2013}, so
\begin{equation}  \label{eqn:mutect_posterior}
  \frac{\mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})}{1 - \mathrm{P}_\mathrm{MuTect}(m,\nu = \hat{f})} = 10^{\mathrm{TLOD} - 6}.
\end{equation}

\subsection*{Implementation}
We have implemented the \batcave algorithm as an R package \texttt{batcaver}.
The package leverages the Bioconductor packages \textrm{BSgenome} \cite{Pages2019}, \textrm{GenomicAlignments} \cite{Lawrence2013}, \textrm{VariantAnnotation} \cite{Obenchain2014}, and \textrm{SomaticSignatures} \cite{Gehring2015} for fast and memory-efficient variant annotation and genomic context identification.
Reference sequences are specified as BSgenome objects, allowing efficient access to genomic context information.

\begin{figure}
\centering
  \includegraphics{figures/sim_signatures_only.pdf}
  \caption{Simulated tumor mutation profiles.
As in Fig.~\ref{NAR-realsigfig}, in each panel the x-axis corresponds to each of the 96 possible mutation types, and the y-axis is the proportion of total mutations of each type.
  (A) A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  (B) Equal proportions of signatures 1, 4, \& 5.
  (C) Equal proportions of signatures 1, 3, \& 5.
   }
  \label{NAR-simsigfig}
 \end{figure}

\subsection*{Tumor simulations}
We used a neutral branching process with no death and $\mu=3\cdot10^{-6}$ to simulate realistic distributions of mutation frequencies.
Tumors were simulated with three different mutation profiles composed of COSMIC mutation signatures (version 2) \cite{Cosmic_consortium}.
Each simulated profile includes COSMIC signature~1, which is found in nearly all tumors and is associated with spontaneous cytosine deamination.
The ``Concentrated'' profile (Fig.~\ref{NAR-simsigfig}A) is an equal combination of COSMIC signatures 1, 7, and 11, which has a large percentage of C $>$ T substitutions such as are often seen in cancers caused by UV exposure \cite{Alexandrov2013}.
The ``Intermediate'' profile (Fig.~\ref{NAR-simsigfig}B) is an equal combination of COSMIC signatures 1, 4, and 5, which has been associated with tobacco carcinogens and is representative of some lung cancers \cite{Alexandrov2013}.
The ``Diffuse'' profile (Fig.~\ref{NAR-simsigfig}C) is an equal combination of COSMIC signatures 1, 3, and 5, which has been associated with inactivating germline mutations in the BRCA1/2 genes leading to a deficiency in DNA double strand break repair \cite{Nik-Zainal2016}.
Simulated variants were sampled from a combination of the Cancer Genome Atlas (TCGA) and Pan-Cancer Analysis of Whole Genomes (PCAWG) databases, which include mutations found in all types of cancer.
Whole genome (100X depth) and whole exome (500X depth) reads were simulated from the GRCh38 reference genome using VarSim \cite{Mu2015} and aligned with BWA \cite{Li2009a}, both with default parameters.
Variants were inserted to create tumors with BAMSurgeon with default parameters~\cite{Ewing2015a} and called with MuTect 1.1.7~\cite{Cibulskis2013} with the following parameters:\\
\texttt{\small java -Xmx24g -jar \$MUTECT\_JAR --analysis\_type MuTect 
        --reference\_sequence \$ref\_path 
        --dbsnp \$db\_snp \\
        --enable\_extended\_output 
        --fraction\_contamination 0.00 
        --tumor\_f\_pretest 0.00 
        --initial\_tumor\_lod -10.00 
        --required\_maximum\_alt\_allele\_mapping\_quality\_score 1 
        --input\_file:normal \$tmp\_normal 
        --input\_file:tumor \$tmp\_tumor 
        --out \$out\_path/\$chr.txt 
        --coverage\_file \$out\_path/\$chr.cov}.\\
Variants identified by MuTect are labelled as to whether they pass all filters, fail to pass only the the evidence threshold \textrm{tlod\_f\_star} filter, or fail to pass any other filter. 
Variants that passed all filters or failed only \textrm{tlod\_f\_star} were then passed to \batcave for prior estimation and rescoring.

\subsection*{Calibration metric}
To quantify the difference in calibration between MuTect and \batcave, we used the Integrated Calibration Index \cite{Austin2019}.
Briefly, a loess-smoothed regression was fit by regressing the binary (True=1, False=0) true variant classification against the reported posterior probability for both MuTect and \batcave.
For a perfectly calibrated caller, the regression fit would be the diagonal line $y=x$. 
The Integrated Calibration Index is a weighted average of the absolute distance between the calibration curve and the diagonal line of perfect calibration.

\subsection*{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

\section*{RESULTS}
We implemented \batcave as a post-call variant evaluation algorithm to be used with MuTect (Versions 1.1.7 or $>$2.0) \cite{Cibulskis2013}.
\batcave extracts the log-likelihood ratio for each potential variant site from the MuTect output, and then it uses that ratio to separate the potential sites into high and low confidence groups.
The mutation profile and mutation rate are estimated from the high confidence sites, and the posterior probability of mutation is then recomputed for all sites.
The \batcave algorithm is inexpensive, processing 22,000 variants per second on a typical desktop computer, which corresponds to roughly 100 seconds to process a 500X exome and 2,000 seconds for a 100X whole genome.

@@

\subsection*{Tests using simulated data}

@@

  \begin{figure*}
\centering
    \includegraphics[width=\textwidth]{figures/fig_wes.pdf}
    \caption{Variant-calling performance on  simulated and real data. Throughout, MuTect results are plotted with gray lines and \batcave results with black lines.
    (A) Precision-recall curves and {(B)} receiver operating characteristic curves for different mutation profiles.
    (C) and {(D)} Calibration plots. Shaded regions show distributions of posterior probabilities for true positive variants, and smooth lines show loess-smoothed relationships, from which the Integrated Calibration Index is calculated. For a perfectly calibrated caller, those curves would match the dashed y=x line. 
    {(E)} and {(F)} Precision-recall curves for real data in which substantial mutation validation was performed \cite{Griffith2015,Shi2018}.
}
  \label{NAR-wes_fig}
  \end{figure*}

\begin{table*}[b]
